### 2005-11-09 doc/EFA/efa.tex
*** empty log message ***

\section{Extended Finite Automaton (EFA)}
An Extended Deterministic Finite Automaton (EFA) is an
augmentation of a regular finite automaton. The regular automaton
$A$ is defined as $A=<Q,\Sigma,\delta, q_i,Q_m>$. Let
$T=\{(q,\sigma)\in Q \times \Sigma: \delta(q,\sigma)!\}$ denote
the set of transitions in the deterministic automaton $A$ and let
$\tau_k$, $k=1,...,|T|$ denote the elements (transitions) of $T$.

The automaton $A$ is extended by a set $X=(x_1,...,x_n)$ of $n$
integer variables $x_{k}$ and by a $|T|$-tuple of function pairs
$G=\big((g_{\tau_1},a_{\tau_1}),...,
(g_{\tau_{|T|}},a_{\tau_{|T|}})\big)$, where each pair
$(g_{\tau_k},a_{\tau_k})$ is associated to the corresponding
transition $\tau_k$ in $T$. $g_{\tau_k}$ is an evaluation of a
boolean formula (a guard formula) over $X$ i.e.
$g_{\tau_k}:{\mathbb{N}}^{n}\rightarrow \{0,1\}$ saying if the
corresponding transition $\tau_k$ is enabled or not. $a_{\tau_k}$
is an updating (action) function from ${\mathbb{N}}^{n}$ to
${\mathbb{N}}^{n}$ of the integer variables $X$.

With the above notations we will formally define an EFA as the
$8$-tuple $A_{e}=<Q, \Sigma, \delta_{e}, q_i, Q_m, X, G>$, where
$\delta_{e}: Q \times \Sigma \times X \rightarrow Q\times X $ is
an extention of the transition function $\delta:Q \times \Sigma
\rightarrow Q$. $\delta_e$ is defined by means of $\delta$ and the
elements of $G$ as
\begin{eqnarray}
\delta_{e}(q,\sigma,X)=\left\{
\begin{array}{ll}
(\delta(q,\sigma),a_{q \sigma}(X)\big)\quad\quad g_{q \sigma}(X)=1\\
\big(q,X \big)\quad\quad\quad\quad g_{q \sigma}(X)=0
\end{array}\right.
\end{eqnarray}
If the guard formula $g_{q \sigma}$ is undefined then
$\delta_{e}(q,\sigma,X)$ is undefined.

\subsection{Synchronous composition of EFA's}
The synchronous composition of two EFA's $A=<Q^A, \Sigma^A,
\delta^A, q^A_i, Q^A_m, X, G^A>$ and $B=<Q^B, \Sigma^B, \delta^B,
q^B_i, Q^B_m, Y, G^B>$ exists if and only if the automata are
consistent. The consistency condition requires that the action
functions triggered by a common enabled  event $\sigma\in
\Sigma_{A}\cap \Sigma_{B}$ in the two automata update common
variables to the same values, i.e. if the $i^{ith}$ variable in
$X$ and the $j^{ith}$ variable in $Y$ are the same variable and
$g_{\alpha \sigma}(X)\wedge g_{\beta \sigma}(Y)=1$ where
$(\alpha,\beta)\in Q^A\times Q^B$ then $(a_{\alpha \sigma}(X))_i =
(a_{\beta \sigma}(Y))_j$.

The consistency condition can be verified by inspection. However,
if the vectors $X$ or $Y$ disable the corresponding guard formula
($g_{\alpha \sigma}(X)\wedge g_{\beta \sigma}(Y)$) it may happen
that $a_{\alpha \sigma}(X))_i \neq (a_{\beta \sigma}(Y))_j$ even
though the two automata are consistent. Therefore it may be a
time-consuming task to verify the consistency condition by
inspection. In section \ref{Implementation} we will present
algorithms that includes a consistency check of the synchronous
composition of two EFA's. \\

Formally we define the synchronous product $A||B$ of two EFA's
$A=<Q^A, \Sigma^A, \delta^A, q^A_i, Q^A_m, X, G^A>$ and $B=<Q^B,
\Sigma^B, \delta^B, q^B_i, Q^B_m, Y, G^B>$ as
\begin{eqnarray}
A||B=<Q^A\times Q^B, \Sigma^A \cup \Sigma^B , \delta, q_i,
Q^A_m\times Q^B_m, X \cup Y, G>.
\end{eqnarray}
To define the guard functions, the action functions and the
transition function for the product we make use of the
corresponding functions in $A$ and $B$.

The guard functions $g_{q \sigma}(Z)=g_{q \sigma}(X\cup Y)$ are
defined as
\begin{eqnarray}
g_{q \sigma}(Z)=\left\{
\begin{array}{ll}
g_{q^A\sigma}(X)\wedge g_{q^B\sigma}(Y) \quad\quad\sigma\in \Sigma^A \cap \Sigma^B\\
g_{q^A\sigma}(X) \quad\quad\sigma\in \Sigma^A \setminus \Sigma^B\\
g_{q^B\sigma}(Y) \quad\quad\sigma\in \Sigma^B \setminus \Sigma^A\\
\textrm{undefined       otherwise}
\end{array}\right.
\end{eqnarray}
If $X\cap Y\neq \emptyset$ we introduce $Y'=Y\setminus X$ and
$X'=X\setminus Y$ and look at the restrictions of the updating
functions $a^A_\delta$ and $a^B_\delta$ to the sets $X'$ and $Y'$.
The updating functions are defined as (if $X\cap Y=\emptyset$ then
$Y'=Y$ and $X'=X$)
\begin{eqnarray}
a_{q \sigma}(Z)=\left\{
\begin{array}{ll}
 \big(a^A_{q^A \sigma}(X), a^B_{q^B \sigma}(Y')\big)\quad\quad\sigma\in \Sigma^A \cap \Sigma^B\\
\big(a^A_{q^A \sigma}(X), Y'\big)\quad\quad\sigma\in \Sigma^A \setminus \Sigma^B\\
\big(X', a^B_{q^B \sigma}(Y)\big)\quad\quad\sigma\in \Sigma^B \setminus \Sigma^A\\
\textrm{undefined       otherwise}
\end{array}\right.
\end{eqnarray}

 The transition function
$\delta(q,\sigma,Z)=\delta(q^A,q^B,\sigma,X\cup Y)$ is defined
from $\delta^A$ and $\delta^B$ as
\begin{eqnarray}
\delta(q,\sigma,Z)=\left\{
\begin{array}{ll}
\big(\delta^A(q^A,\sigma),\delta^B(q^B,\sigma),a_{q \sigma}(Z)\big) \quad\quad \sigma\in \Sigma^A \cap \Sigma^B \quad g_{q \sigma}(Z)=1\\
\big(\delta^A(q^A,\sigma),q^B,a_{q \sigma}(Z)\big) \quad\quad \sigma\in \Sigma^A \setminus \Sigma^B \quad g_{q \sigma}(Z)=1\\
\big(q^A,\delta^B(q^B,\sigma),a_{q \sigma}(Z)\big) \quad\quad \sigma\in \Sigma^B \setminus \Sigma^A \quad g_{q \sigma}(Z)=1\\
\textrm{undefined       otherwise}
\end{array}\right.
\end{eqnarray}

\noindent By the above definition of synchronization we now obtain
the following property

@@

\noindent \textbf{Proof}: We show this by considering three
automata $A, A'$ and $A''$ and showing that $(A\|A')\|A'' =
A\|(A'\|A'')$.

The 8-tuple representing $(A\|A')\|A''$ is

\begin{center}
\begin{tabular}{ll}
  $(A\|A')\|A'' =$ & $\Big \langle (Q \times Q') \times Q'',$\\
  & $ \big(\Sigma \bigcup \Sigma' \big) \bigcup \Sigma'',$\\
  & $\delta^{(A\|A')\|A''},$\\
  & $\big((q_i, q'_i) , q''_i\big),$\\
  & $(Q_m \times Q_m') \times Q_m'',$\\
  & $\big(X \bigcup X') \bigcup X'',$\\
  & $G^{(A\|A')\|A''},$\\
  & $\big(X_m \bigcup X'_m) \bigcup X''_m \Big \rangle$
\end{tabular}
\end{center}

@@

\noindent Now consider the action function $a^{(A\|A')\|A''} : X^c
\rightarrow X^c$. Here things get a bit more messy, since we need
to make sure that each variable is updated at most one time even
though we might have $X\bigcap X'\bigcap X'' \neq \emptyset$

>>>>

\section{Extended Finite Automaton (EFA)}
An Extended Deterministic Finite Automaton (EFA) is an
augmentation of a regular finite automaton. The regular automaton
$A$ is defined as $A=\langle Q,\Sigma,\delta, q_i,Q_m \rangle$.
Let $T=\{(q,\sigma)\in Q \times \Sigma: \delta(q,\sigma)!\}$
denote the set of transitions in the deterministic automaton $A$
and let $\tau_k$, $k=1,...,|T|$ denote the elements (transitions)
of $T$.

The automaton $A$ is extended by a vector $X=(x_1,...,x_n)$ of $n$
finite integer variables $x_{k}$ and by a $|T|$-tuple of function
pairs $G=\big((g_{\tau_1},a_{\tau_1}),...,
(g_{\tau_{|T|}},a_{\tau_{|T|}})\big)$, where each pair
$(g_{\tau_k},a_{\tau_k})$ is associated to the corresponding
transition $\tau_k$ in $T$. $g_{\tau_k}$ is an evaluation of a
boolean formula (a guard formula) of $X$ i.e.
$g_{\tau_k}:{\mathbb{N}}^{n}\rightarrow \{0,1\}$ saying if the
corresponding transition $\tau_k$ is enabled or not. $a_{\tau_k}$
is an updating (action) function from ${\mathbb{N}}^{n}$ to
${\mathbb{N}}^{n}$ of the integer variables $X=(x_1,...,x_n)$.

With the above notations we will formally define an EFA as the
$8$-tuple $A_{e}=\langle Q, \Sigma, \delta_{e}, q_i, Q_m, X, G
,X_m \rangle$, where $X_m$ are the marked integer values of the
variables in $X$ and $\delta_{e}: Q \times \Sigma \times
{\mathbb{N}}^{n} \rightarrow Q\times {\mathbb{N}}^{n} $ is a new
extended version of the transition function $\delta:Q \times
\Sigma \rightarrow Q$. The extended transition function $\delta_e$
is defined by means of $\delta$ and the elements of $G$ as
\begin{eqnarray}
\delta_{e}(q,\sigma,X)=\left\{
\begin{array}{ll}
(\delta(q,\sigma),a_{q \sigma}(X)\big)\quad\quad g_{q \sigma}(X)=1\\
\textrm{undefined} \quad\quad\quad\quad\quad g_{q \sigma}(X)=0
\end{array}\right.
\end{eqnarray}
If $(q,\sigma)\notin T$ then $g_{q\sigma}$ is undefined and
therefore $\delta_{e}(q,\sigma,X)$ is undefined. Notice that for
the set of transitions $T_e$ in the extended automaton $A_e$ we
have $T_e=\{(q,\sigma)\in Q \times \Sigma: g_{q\sigma}(X)=1
\textrm{ for some } X\}\subset \{(q,\sigma)\in Q \times \Sigma:
g_{q\sigma} \textrm{ is defined } \}\ =T$.

\subsection{Synchronous composition of EFA's}
The synchronous composition of two EFA's $A=\langle Q^A, \Sigma^A,
\delta^A, q^A_i, Q^A_m, X^{A}, G^A, X^A_m \rangle$ and $B=\langle
Q^B, \Sigma^B, \delta^B, q^B_i, Q^B_m, X^{B}, G^B, X^B_m \rangle$
exists if and only if the automata are consistent. The consistency
condition requires that the action functions triggered by a common
enabled event $\sigma\in \Sigma_{A}\cap \Sigma_{B}$ in the two
automata update common variables to the same values, i.e. if the
$i^{ith}$ variable in $X^{A}$ and the $j^{ith}$ variable in
$X^{B}$ are the same variable and $g_{q^A \sigma}(X^{A})\wedge
g_{q^B \sigma}(X^{B})=1$ where $(q^A,q^B)\in Q^A\times Q^B$ then
$(a_{q^A \sigma}(X^{A}))_i = (a_{q^B \sigma}(X^{B}))_j$.

The consistency condition can be verified by inspection. However,
if the vectors $X^{A}$ or $X^{B}$ disable the corresponding guard
formula ($g_{q^A \sigma}(X^{A})\wedge g_{q^B \sigma}(X^{B})$) it
may happen that $a_{q^A \sigma}(X^{A}))_i \neq (a_{q^B
\sigma}(X^{B}))_j$ even though the two automata are consistent.
Therefore it may be a time-consuming task to verify the
consistency condition by inspection. In section
\ref{Implementation} we will present algorithms that includes a
consistency check of the synchronous
composition of two EFA's. \\

 Formally we define the synchronous product $A||B$ of two EFA's
$A=\langle Q^A, \Sigma^A, \delta^A, q^A_i, Q^A_m, X^{A}, G^A,X^A_m
\rangle$ and $B=\langle Q^B, \Sigma^B, \delta^B, q^B_i, Q^B_m,
X^{B}, G^B,X^B_m
 \rangle$ as
\begin{eqnarray}
A||B=\langle Q^A\times Q^B, \Sigma^A \cup \Sigma^B , \delta, q_i,
Q^A_m\times Q^B_m, Z, G, Z_m \rangle.
\end{eqnarray}
To define the integer variables, the guard functions, the action
functions and the transition function for the product we make use
of the corresponding elements in $A$ and $B$.

 Let $Z^{AB}$ denote the common variables of $X^{A}$ and $X^{B}$, and let $Z^A$
and $Z^B$ be the distinct variables of $X^{A}$ and $X^{B}$,
respectively. We define the integer variables $Z$ of the
synchronous product of $A$ and $B$ as the vector $Z=(Z^A, Z^{AB},
Z^B)$. The guard functions $g_{q \sigma}(Z)$ are defined as
\begin{eqnarray}\label{guard_sync_1}
g_{q \sigma}(Z)=\left\{
\begin{array}{ll}
g_{q^A\sigma}(X^{A})\wedge g_{q^B\sigma}(X^{B}) \textrm{ if }\sigma\in \Sigma^A \cap \Sigma^B\\
g_{q^A\sigma}(X^{A}) \quad\quad \textrm{ if } \sigma\in \Sigma^A \setminus \Sigma^B\\
g_{q^B\sigma}(X^{B}) \quad\quad \textrm{ if }
\sigma\in\Sigma^B\setminus \Sigma^A
\end{array}\right.
\end{eqnarray}
If any one of the guards functions in the above definition are
undefined the guard function for the product $g_{q \sigma}$ is
also undefined. To simplify the notation further, we use "dummy"
guard functions in the following way. If $\sigma\in \Sigma^A
\setminus \Sigma^B$ we introduce fictional guards functions
$g_{q^B\sigma}(.)=1$ for $B$ and if $\sigma\in\Sigma^B\setminus
\Sigma^A$ we set $g_{q^A\sigma}(.)=1$. With this convention we can
write equation (\ref{guard_sync_1}) as
\begin{eqnarray}
g_{q \sigma}(Z)= g_{q^A\sigma}(X^{A})\wedge g_{q^B\sigma}(X^{B}).
\end{eqnarray}
To define the action functions for $A||B$ we look at the
restrictions of the updating functions $a^A_\delta$ and
$a^B_\delta$ to the sets $Z^A$ and $Z^B$. The updating functions
are defined as (if $X^{A}$ and $X^{B}$ have no common variables
then $Z^A=X^{A}$ and $Z^B=X^{B}$)
\begin{eqnarray}
a_{q \sigma}(Z)=\left\{
\begin{array}{ll}
 \big(a^A_{q^A \sigma}(X^{A}), a^B_{q^B \sigma}(Z^{B})\big)\quad\quad\sigma\in \Sigma^A \cap \Sigma^B\\
\big(a^A_{q^A \sigma}(X^{A}), Z^{B}\big)\quad\quad\quad\sigma\in \Sigma^A \setminus \Sigma^B\\
\big(Z^{A}, a^B_{q^B \sigma}(X^{B})\big)\quad\quad\quad\sigma\in
\Sigma^B \setminus \Sigma^A
\end{array}\right.
\end{eqnarray}
As in the definition of the guard functions the notion of being
undefined is inherited from the action functions of $A$ and $B$.
Similar to the "dummy" guard functions we introduce fictional
action functions. If $\sigma\in \Sigma^A \setminus \Sigma^B$ we
introduce fictional action functions $a_{q^B\sigma}(X^B)=X^B$ for
$B$ and if $\sigma\in\Sigma^B\setminus \Sigma^A$ we set
$a_{q^A\sigma}(X^A)=X^A$. If $X^{A}$ and $X^{B}$ have no common
variables then the formula for the action functions of the product
simplifies to
\begin{eqnarray}
a_{q \sigma}(Z)=
 \big(a^A_{q^A \sigma}(X^{A}), a^B_{q^B \sigma}(X^{B})\big).
\end{eqnarray}

 The transition function
$\delta(q,\sigma,Z)$ is defined from $\delta^A$ and $\delta^B$ as
\begin{eqnarray}
\delta(q,\sigma,Z)=\left\{
\begin{array}{ll}
\big(\delta^A(q^A,\sigma),\delta^B(q^B,\sigma),a_{q \sigma}(Z)\big) \quad\quad \sigma\in \Sigma^A \cap \Sigma^B \quad g_{q \sigma}(Z)=1\\
\big(\delta^A(q^A,\sigma),q^B,a_{q \sigma}(Z)\big) \quad\quad\quad\quad\quad \sigma\in \Sigma^A \setminus \Sigma^B \quad g_{q \sigma}(Z)=1\\
\big(q^A,\delta^B(q^B,\sigma),a_{q \sigma}(Z)\big) \quad\quad\quad\quad\quad \sigma\in \Sigma^B \setminus \Sigma^A \quad g_{q \sigma}(Z)=1\\
\textrm{
undefined}\quad\quad\quad\quad\quad\quad\quad\quad\quad\quad\quad\quad\quad\quad\quad\quad\quad
g_{q\sigma}(Z)=0
\end{array}\right.
\end{eqnarray}

The synchronous product $A||B$ of two FSA $A$ and $B$ can be
extended to $(A||B)_e$ by taking the synchronous product of the
extended components  $A_e$ and $B_e$. In other words, the
operations extension and product commute, $(A||B)_e=A_e||B_e$.

Moreover, by the above definition of synchronization we now obtain
the following property

@@

\noindent \textbf{Proof}: We show this by considering three
automata $A, A'$ and $A''$ and showing that $(A\|A')\|A'' =
A\|(A'\|A'')$.
\newline\newline
\noindent The 8-tuple representing $(A\|A')\|A''$ is
\begin{center}
\begin{tabular}{ll}
  $(A\|A')\|A'' =$ & $\Big \langle (Q \times Q') \times Q'',$\\
  & $ \big(\Sigma \bigcup \Sigma' \big) \bigcup \Sigma'',$\\
  & $\delta^{(A\|A')\|A''},$\\
  & $\big((q_i, q'_i) , q''_i\big),$\\
  & $(Q_m \times Q_m') \times Q_m'',$\\
  & $\big[[X, X'], X'']$\\
  & $G^{(A\|A')\|A''},$\\
  & $\big[[X_m, X'_m], X''_m \big] \Big \rangle$
\end{tabular}
\end{center}

@@

\noindent Now consider the action function $a^{(A\|A')\|A''} : X^c
\rightarrow X^c$.
